---
title: 'Problem 4: Largest palindrome product'
author: John Ensley
date: '2018-06-29'
slug: '004'
categories: []
tags: []
weight: -4
hidedate: true
---

# [Problem](https://projecteuler.net/problem=4)

> A palindromic number reads the same both ways. The largest palindrome made from the product of two 2-digit numbers is 9009 = 91 × 99.
> 
> Find the largest palindrome made from the product of two 3-digit numbers.

# Explanation

The overall approach for this problem is very simple: we'll loop through all combinations of two three-digit numbers, multiply them, and see if the product is a palindrome or not. If it is a palindrome, and it's bigger than any palindrome we've seen so far, then we make a note of it. At the end we'll be left with the largest palindrome product of 3-digit numbers.

# Implementation

There are 900 three-digit numbers, so iterating through all combinations would require $900*900 = 810,000$ calls to some palindrome-checking function. This could potentially take a long time. For efficiency, we should consider two things. First, we should try to cut down on the number of times we need to check if a number is a palindrome. Second, we should be smart about how we check for palindrome-ness.

Let's tackle the palindrome function first. There are a few ways we could check whether a number $x$ is a palindrome. This function converts $x$ to a vector of digits and then compares it to the reversed version of itself.

```{r}
is_palindrome1 <- function(x) {
  digit_vector <- as.numeric(strsplit(as.character(x), split = '')[[1]])
  all(digit_vector == rev(digit_vector))
}
```

This one builds up a vector of digits using modular arithmetic instead of converting to a string and back to a number as in `is_palindrome1`. It again compares the vector to its reverse.

```{r}
is_palindrome2 <- function(x) {
  ndigits <- floor(log10(x)) + 1
  digits <- rep(NA, ndigits)
  for (i in rev(seq_along(digits))) {
    digits[i] <- x %% 10
    x <- x %/% 10
  }
  all(digits == rev(digits))
}
```

Instead of building up a vector of digits, this one builds up a reversed integer. For example, if $x = 1234$, `reversed` will become $4321$. Then $x$ and `reversed` are compared.

```{r}
is_palindrome3 <- function(x) {
  n <- x
  reversed <- 0
  while (x > 0) {
    remainder <- x %% 10
    reversed <- 10 * reversed + remainder
    x <- x %/% 10
  }
  n == reversed
}
```

In this one, we aren't reversing $x$, but instead we compare the first and last digits of $x$. If they are not equal, then we already know $x$ is not a palindrome and we can exit the function without checking the rest. If they are equal, we remove the first and last digit from $x$ and perform the comparison again on this new $x$.

```{r}
is_palindrome4 <- function(x) {
  d <- 10^(floor(log10(x)))
  r <- 10
  while (x > 0) {
    dig_first <- x %/% d
    dig_last <- x %% r
    if (dig_first != dig_last) return(FALSE)
    x <- x %% d %/% r
    d <- d / 100
  }
  return(TRUE)
}
```

`is_palindrome4` has the advantage of potentially exiting early, although there are many more modular arithmetic calculations needed than in `is_palindrome3`. I'm not sure how fast R can perform numeric to string conversions, so I don't know how well `is_palindrome1` will perform compared to `is_palindrome2`. Let's find out! First let's test all four functions on a palindromic number. Since products of 3-digit numbers must be 5 or 6 digits, we won't worry about testing very large numbers.

```{r}
(m_pal <- microbenchmark::microbenchmark(is_palindrome1(456654),
                                         is_palindrome2(456654),
                                         is_palindrome3(456654),
                                         is_palindrome4(456654),
                                         unit = 'relative',
                                         control = list(warmup = 10)))
```

Note that the reported values are relative to the fastest runtime, not absolute speed. `r tufte::margin_note('The runtimes for all four functions are only a few microseconds, so the actual values really aren\'t that important.')` `is_palindrome3` and `is_palindrome4` perform just about equally well, and the other two are 2 to 3 times slower. For the sake of completeness, let's try testing on a non-palindrome as well.

```{r}
(m_pal <- microbenchmark::microbenchmark(is_palindrome1(468841),
                                         is_palindrome2(468841),
                                         is_palindrome3(468841),
                                         is_palindrome4(468841),
                                         unit = 'relative',
                                         control = list(warmup = 10)))
```

As we would expect, `is_palindrome4` performs the best because it quits as soon as it encounters digits that do not match. This will probably come in handy, since there are more non-palindromes than palindromes. Let's go with `is_palindrome4` as our final answer.



The next step is to find the largest palindrome product. Again, the general idea is to iterate through all products $ab$ where $a$ and $b$ are three-digit numbers. There are a couple of optimizations that minimize the number of times we'll need to call `is_palindrome4`. First of all, because the largest palindrome will likely be the product of two larger three-digit numbers, we can start with $a$ and $b$ equal to 999 and move backwards to 100. That way we'll probably encounter the largest palindrome sooner.

Second, and more importantly, we don't need to check whether every single $ab$ is a palindrome. If $ab$ is less than the largest palindrome we've found so far, then we don't care if it is also a palindrome.

```{r}
p4 <- function() {
  largest <- -Inf
  for (a in 999:100) {
    for (b in a:100) {
      x <- a*b
      if (x > largest && is_palindrome4(x)) {
        largest <- x
      }
    }
  }
  largest
}
```

```{r}
p4()
```


# Timing

```{r}
( m <- microbenchmark::microbenchmark(p4(), unit = 'ms') )
```

The solution takes about `r round(microbenchmark:::summary.microbenchmark(m)$median, 1)` milliseconds. If the predicates in the `if` statement were reversed, i.e. `if (is_palindrome4(x) && x > largest)`, it would take about 2 seconds, or roughly 50 times longer.
