---
title: 'Problem 5: Smallest multiple'
author: John Ensley
date: '2018-07-03'
slug: '005'
categories: []
tags: [gcd, lcm]
weight: -5
hidedate: true
draft: false
---

#[Problem](https://projecteuler.net/problem=5)

> 2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder.
> 
> What is the smallest positive number that is evenly divisible by all of the numbers from 1 to 20?

# Explanation

This problem boils down to finding the [least common multiple](https://en.wikipedia.org/wiki/Least_common_multiple) of all numbers from 1 to 20. `r tufte::margin_note('The least common multiple, or LCM, of two integers $a$ and $b$ is the smallest positive integer that is divisible by both $a$ and $b$.')` There are many ways to calculate the LCM for a set of integers, but one of the simpler ones is to write it in terms of the [greatest common divisor](https://en.wikipedia.org/wiki/Greatest_common_divisor), or GCD. `r tufte::margin_note('The greatest common divisor, or GCD, of two integers $a$ and $b$ is the largest positive integer that divides both $a$ and $b$.')`

The relationship between the LCM and the GCD is as follows:

$$
LCM(a, b) = \frac{|a \cdot b|}{GCD(a, b)}.
$$

The GCD can be calculated very efficiently using [Euclid's algorithm](https://en.wikipedia.org/wiki/Euclidean_algorithm). This algorithm takes advantage of the fact that the GCD doesn't change if you replace $b$ with $b \; \textrm{mod} \; a$. `r tufte::margin_note('Or vice versa if $a$ is larger than $b$.')`

# Implementation

First, we implement the GCD function.

```{r}
gcd <- function(a, b) {
  while(b != 0) {
    t <- b
    b <- a %% b
    a <- t
  }
  a
}

gcd(6, 27)
```

Then the LCM is very simple:

```{r}
lcm <- function(a, b) {
  a / gcd(a, b) * b
}
lcm(6, 27)
```

This works for the LCM of two values, but the relationship between LCM and GCD does not hold for three or more values. We need to find the LCM of all integers from 1 to 20. To do this, we can use the fact that

$$
GCD(a, b, c) = GCD(a, GCD(b, c)).
$$

By iteratively finding the GCD of each pair of integers, we will arrive at the overall GCD.

```{r}
lcm <- function(v) {
  n <- length(v)
  if (n < 2) stop('Must supply a vector of at least 2 integers')
  
  x <- v[1]
  for (i in 2:n) {
    x <- x / gcd(x, v[i]) * v[i]
  }
  x
}
```

The answer is:

```{r}
lcm(1:20)
```

# Timing


```{r}
(m <- microbenchmark::microbenchmark(lcm(1:20), unit = 'ms'))
```

The solution takes about `r round(microbenchmark:::summary.microbenchmark(m)$median, 2)` milliseconds.