---
title: 'Problem 5: Smallest multiple'
author: John Ensley
date: '2018-07-03'
slug: '005'
categories: []
tags: [gcd, lcm]
weight: -5
hidedate: true
---

#[Problem](https://projecteuler.net/problem=5)

> 2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder.
> 
> What is the smallest positive number that is evenly divisible by all of the numbers from 1 to 20?

# Explanation

This problem boils down to finding the [least common multiple](https://en.wikipedia.org/wiki/Least_common_multiple) of all numbers from 1 to 20. `r tufte::margin_note('The least common multiple, or LCM, of two integers $a$ and $b$ is the smallest positive integer that is divisible by both $a$ and $b$.')` There are many ways to calculate the LCM for a set of integers, but one of the simpler ones is to write it in terms of the [greatest common divisor](https://en.wikipedia.org/wiki/Greatest_common_divisor), or GCD. `r tufte::margin_note('The greatest common divisor, or GCD, of two integers $a$ and $b$ is the largest positive integer that divides both $a$ and $b$.')`

The relationship between the LCM and the GCD is as follows:

$$
LCM(a, b) = \frac{|a \cdot b|}{GCD(a, b)}.
$$

The GCD can be calculated very efficiently using [Euclid's algorithm](https://en.wikipedia.org/wiki/Euclidean_algorithm). This algorithm takes advantage of the fact that the GCD doesn't change if you replace $b$ with $b \; \textrm{mod} \; a$. `r tufte::margin_note('Or, if $a$ is larger than $b$, replace $a$ with $a \; \textrm{mod} \; b$.')`

# Implementation


```{r}
gcd <- function(a, b) {
  while(b != 0) {
    t <- b
    b <- a %% b
    a <- t
  }
  a
}
```



```{r}
lcm <- function(a, b) {
  a / gcd(a, b) * b
}
```

```{r}
lcm <- function(v) {
  n <- length(v)
  if (n < 2) stop('Must supply a vector of at least 2 integers')
  
  x <- v[1]
  for (i in 2:n) {
    x <- x / gcd(x, v[i]) * v[i]
  }
  x
}
```


```{Rcpp eval=FALSE, include=FALSE}
# include <Rcpp.h>
using namespace Rcpp;

int gcd_cpp(int a, int b)
{
  if (b == 0) return a;
  else return gcd_cpp(b, a % b);
}

// [[Rcpp::export]]
int lcm_cpp(NumericVector v)
{
  int n = v.size();
  int x = v[0];
  for (int i = 1; i < n; ++i)
  {
    x = x / gcd_cpp(x, v[i]) * v[i];
  }
  return x;
}
```




# Timing


```{r}
microbenchmark::microbenchmark(lcm(1:20))
```

