---
title: 'Problem 6: Sum square difference'
author: John Ensley
date: '2018-07-06'
slug: 6
categories: []
tags:
  - sum of squares
weight: -6
hidedate: true
draft: false
---

# [Problem](https://projecteuler.net/problem=6)

> The sum of the squares of the first ten natural numbers is,

$$
1^2 + 2^2 + \dots + 10^2 = 385
$$

> The square of the sum of the first ten natural numbers is,

$$
(1 + 2 + ... + 10)^2 = 55^2 = 3025
$$

> Hence the difference between the sum of the squares of the first ten natural numbers and the square of the sum is 3025 − 385 = 2640.
>
> Find the difference between the sum of the squares of the first one hundred natural numbers and the square of the sum.

# Explanation

This one is trivially easy. Just do the calculation. You could be clever about using the fact that

$$
\sum_{i=1}^n x_i = \frac{n(n+1)}{2}
$$

and other such relations, but just calculating the answer using `sum()` is so fast that it's not even worth it.

# Implementation

```{r}
p6 <- function(n) {
  integers <- 1:n
  sum(integers)^2 - sum(integers^2)
}

p6(100)
```

# Timing

```{r}
(m <- microbenchmark::microbenchmark(p6(100), unit = 'ms'))
```

The solution takes about `r round(microbenchmark:::summary.microbenchmark(m)$median, 3)` milliseconds.