---
title: 'Problem 7: 10001st prime'
author: John Ensley
date: '2018-07-12'
slug: '7'
categories: []
tags: [sieve of Eratosthenes, primes]
weight: -7
hidedate: true
draft: true
---

# [Problem](https://projecteuler.net/problem=7)

> By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.
>
> What is the 10 001st prime number?

# Explanation

We've already implemented the [sieve of Eratosthenes](/3/), a method for generating all primes less than some $n$. In order to use that, we'll need to know some value of $n$ such that there will be at least 10,001 prime numbers less than $n$. As it turns out, [J. Barkley Rosser](https://www.jstor.org/stable/2371291) established just such an upper bound. For the $n$th prime $p_n$,

$$
p_n < n \ln (n \ln n).
$$

Therefore, $p_{10001} < 10001 \ln (10001 \ln 10001) = 114319.2 < 114320$.

# Implementation

```{r}
sieve <- function(n) {
  is_prime <- c(F, rep(T, n-1))
  for (i in 2:sqrt(n)) {
    if (is_prime[i]) {
      is_prime[seq(i*i, n, by = i)] <- F
    }
  }
  which(is_prime)
}
```


```{r}
p7 <- function(n) {
  limit <- n * log(n * log(n))
  primes <- sieve(limit)
  primes[n]
}
```

```{r}
p7(10001)
```

# Timing

```{r}
( m <- microbenchmark::microbenchmark(p7(10001), unit='ms') )
```

The solution takes about `r round(microbenchmark:::summary.microbenchmark(m)$median, 1)` milliseconds.